---
title: "Module Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Module Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datfloish)
```

## Overview

datfloish is built from three Shiny modules, each responsible for a single
slice of the application. The modules are wired together in `app_server()`,
which threads a shared `db_path` and `interval_ms` through the pieces that
need them.

```
app_ui()                          app_server()
├── mod_add_sale_ui()             ├── mod_add_sale_server()     ─── writes to db
├── mod_value_boxes_ui()          ├── mod_sales_table_server()  ─── polls db, returns stats
└── mod_sales_table_ui()          └── mod_value_boxes_server()  ─── renders stats
```

The only cross-module communication is a named list of `reactiveVal`s
(`poll_stats`) that `mod_sales_table_server()` returns and
`mod_value_boxes_server()` consumes. Everything else is isolated inside its
own module namespace.

## The three modules

### `mod_add_sale` — writing to the database

This module owns the sidebar form. Its server function listens for the
"Add Sale" button and inserts one row into the `sales` table on each click:

```{r add-sale-server, eval = FALSE}
mod_add_sale_server <- function(id, db_path) {
  moduleServer(id, function(input, output, session) {
    observeEvent(input$add, {
      con <- dbConnect(SQLite(), db_path)
      on.exit(dbDisconnect(con))
      dbExecute(
        con,
        "INSERT INTO sales (product, amount, updated_at) VALUES (?, ?, ?)",
        params = list(input$product, input$amount,
                      format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
      )
    })
  })
}
```

The module does not return a value; it acts purely through the side effect of
writing to the database. The polling module detects that write on its next
`checkFunc` tick.

### `mod_sales_table` — polling and displaying

This is the core module. Its server function creates a `reactivePoll()` that:

1. Runs `checkFunc` every `interval_ms` milliseconds — a lightweight
   `SELECT MAX(updated_at)` query.
2. Runs `valueFunc` only when `checkFunc` returns a new value — the full
   `SELECT *` that feeds the table.

Alongside the poll, it tracks four `reactiveVal`s and returns them so other
modules can display them:

```{r sales-table-return, eval = FALSE}
list(
  check_count = check_count,   # number of checkFunc calls
  fetch_count = fetch_count,   # number of valueFunc calls
  last_check  = last_check,    # timestamp of most recent check
  last_fetch  = last_fetch     # timestamp of most recent fetch
)
```

The table output (`output$sales_table`) is rendered inside the module — no
data leaves the namespace.

### `mod_value_boxes` — displaying poll statistics

This module receives the `reactiveVal` list from `mod_sales_table_server()`
and renders it into four `bslib::value_box()` elements. It holds no state
of its own and performs no database access:

```{r value-boxes-server, eval = FALSE}
mod_value_boxes_server <- function(id, stats) {
  moduleServer(id, function(input, output, session) {
    output$check_count <- renderText(as.character(stats$check_count()))
    output$fetch_count <- renderText(as.character(stats$fetch_count()))
    output$last_check  <- renderText(stats$last_check())
    output$last_fetch  <- renderText(stats$last_fetch())
  })
}
```

Keeping the display logic separate from the polling logic means the value
boxes can be restyled or replaced without touching the polling code.

## Wiring in `app_server()`

`app_server()` is the only place where the three modules are aware of each
other. It is deliberately thin:

```{r app-server, eval = FALSE}
app_server <- function(
    db_path     = system.file("extdata", "sales.db", package = "datfloish"),
    interval_ms = 5000) {
  function(input, output, session) {

    # mod_sales_table_server returns the reactiveVal list
    poll_stats <- mod_sales_table_server("sales",
      db_path     = db_path,
      interval_ms = interval_ms
    )

    # db_path only; does not need poll_stats
    mod_add_sale_server("add_sale", db_path = db_path)

    # receives poll_stats; does not need db_path
    mod_value_boxes_server("stats", stats = poll_stats)
  }
}
```

`db_path` flows into two modules independently — the add-sale module and the
sales-table module each open and close their own connections as needed. There
is no shared connection object, which avoids concurrency issues with SQLite.

## Data flow

The diagram below shows how information moves through the application on a
typical user interaction (adding a sale):

```
User clicks "Add Sale"
        │
        ▼
mod_add_sale_server()
  INSERT INTO sales ...
        │
        │  (database write)
        ▼
mod_sales_table_server() — checkFunc tick
  SELECT MAX(updated_at)  ← new value detected
        │
        ▼
mod_sales_table_server() — valueFunc fires
  SELECT * FROM sales
  fetch_count + 1
  last_fetch updated
        │
        ├──► output$sales_table   (table re-renders)
        │
        └──► poll_stats (reactiveVals)
                  │
                  ▼
        mod_value_boxes_server()
          Fetches counter increments
          Last Fetch timestamp updates
```

Between sales, only `checkFunc` runs. The value boxes, table, and
`fetch_count` remain unchanged until the next write reaches the database.

## The `db_path` convention

Every function that touches the database accepts a `db_path` argument and
defaults to the bundled database:

```{r db-path, eval = FALSE}
db_path = system.file("extdata", "sales.db", package = "datfloish")
```

This means `launch_app()` works out of the box with no setup. Passing a
custom path — such as a `tempfile()` — is the standard pattern in tests and
when running multiple isolated app instances.

```{r custom-db, eval = FALSE}
my_db <- tempfile(fileext = ".db")
seed_db(db_path = my_db)
launch_app(db_path = my_db)
```
