---
title: "Efficient Database Polling with reactivePoll()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efficient Database Polling with reactivePoll()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datfloish)
```

## The polling problem

A live dashboard needs to periodically check whether its data source has
changed. The simplest approach is to re-query the database on a fixed timer
and update every output regardless of whether anything is new:

```{r naive, eval = FALSE}
# Naive approach: re-query on every tick
sales_data <- reactive({
  invalidateLater(5000)
  con <- dbConnect(SQLite(), db_path)
  on.exit(dbDisconnect(con))
  dbGetQuery(con, "SELECT * FROM sales ORDER BY updated_at DESC")
})
```

This works, but it conflates two separate concerns:

1. **Has the data changed?** — a cheap question answered by a single scalar.
2. **What is the new data?** — an expensive question that transfers every row.

When nothing has changed, the full query runs anyway, wasting database I/O,
network bandwidth (for remote databases), and serialisation time every single
tick.

## Separating check from fetch

`reactivePoll()` solves this by accepting two functions instead of one:

| Argument | Purpose | Ideal cost |
|---|---|---|
| `checkFunc` | Returns a scalar that changes when data changes | Very low |
| `valueFunc` | Returns the full result set | Higher |

Shiny runs `checkFunc` on every `intervalMillis` tick. `valueFunc` only
runs when the value returned by `checkFunc` differs from the previous call.
If nothing has changed, `valueFunc` is never invoked.

```{r reactive-poll-sketch, eval = FALSE}
reactivePoll(
  intervalMillis = 5000,
  session        = session,

  checkFunc = function() {
    con <- dbConnect(SQLite(), db_path)
    on.exit(dbDisconnect(con))
    dbGetQuery(con, "SELECT MAX(updated_at) FROM sales")[[1]]  # one value
  },

  valueFunc = function() {
    con <- dbConnect(SQLite(), db_path)
    on.exit(dbDisconnect(con))
    dbGetQuery(con, "SELECT * FROM sales ORDER BY updated_at DESC")  # full table
  }
)
```

The database only bears the cost of a full table scan when a new record has
actually been written.

## How datfloish implements this

`mod_sales_table_server()` wraps the pattern above and exposes four
`reactiveVal`s so that the rest of the UI can visualise the difference in
real time:

```{r mod-server-excerpt, eval = FALSE}
check_count <- reactiveVal(0)
fetch_count <- reactiveVal(0)

sales_data <- reactivePoll(
  intervalMillis = interval_ms,
  session        = session,

  checkFunc = function() {
    check_count(isolate(check_count()) + 1)      # increments every tick
    last_check(format(Sys.time(), "%H:%M:%S"))

    con <- dbConnect(SQLite(), db_path)
    on.exit(dbDisconnect(con))
    if (!dbExistsTable(con, "sales")) return(NA)
    dbGetQuery(con, "SELECT MAX(updated_at) FROM sales")[[1]]
  },

  valueFunc = function() {
    fetch_count(isolate(fetch_count()) + 1)      # increments only on change
    last_fetch(format(Sys.time(), "%H:%M:%S"))

    con <- dbConnect(SQLite(), db_path)
    on.exit(dbDisconnect(con))
    dbGetQuery(con, "SELECT * FROM sales ORDER BY updated_at DESC")
  }
)
```

## Observing the difference

Launch the app, wait a minute without adding any sales, and watch the value
boxes:

- **Checks Performed** increments every 5 seconds — one lightweight
  `SELECT MAX(updated_at)` per tick.
- **Fetches Performed** stays fixed — `valueFunc` has nothing new to return.

Now click **Add Sale**. The next `checkFunc` call returns a new
`MAX(updated_at)`, Shiny detects the change, and `valueFunc` fires exactly
once. The Fetches counter increments by one; the table updates.

```{r launch, eval = FALSE}
launch_app()
```

After running the app for two minutes with five sales added, you might see:

| Metric | Value |
|---|---|
| Checks Performed | 24 |
| Fetches Performed | 5 |

Twenty-four cheap timestamp reads; five full-table fetches. Without
`reactivePoll()` the same period would have required twenty-four full-table
reads instead.

## When the savings matter most

The efficiency gap widens as:

- **Rows increase** — a full `SELECT *` becomes more expensive.
- **Poll interval decreases** — more ticks per minute means more skipped
  fetches.
- **Write frequency is low relative to polling frequency** — the common case
  for dashboards where data changes infrequently but must appear promptly.
- **The database is remote** — network latency and connection overhead make
  each avoided round-trip meaningful.

For a local SQLite database with ten rows, the savings are negligible. For a
PostgreSQL table with millions of rows polled every second, they are
substantial.
